%
% Copyright © 2020 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../latex/blogpost.tex}
\renewcommand{\basename}{fibonacci}
%\renewcommand{\dirname}{notes/phy1520/}
\renewcommand{\dirname}{notes/ece1228-electromagnetic-theory/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../latex/peeter_prologue_print2.tex}

\usepackage{mathtools}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}

\usepackage{amsthm}
\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}
\usepackage{verbatim}
%\usepackage{mhchem} % \ce{}
%\usepackage{macros_bm} % \bcM
%\usepackage{macros_qed} % \qedmarker
%\usepackage{txfonts} % \ointclockwise
%\usepackage{amsmath} % \binom

%\renewcommand{\binom}[2]{{{#1}\choose{#2}}}

\beginArtNoToc

\generatetitle{The nth term of a Fibonacci series.}
%\chapter{The nth term of a Fibonacci series.}
%\label{chap:fibonacci}

I've just started reading \citep{strogatz2009calculus}, but already got distracted from the plot by a fun math fact.  Namely, a cute formula for the nth term of a Fibonacci series.  Recall
\makedefinition{Fibonacci series.}{dfn:fibonacci:20}{
With \( F_0 = 0 \), and \( F_1 = 1 \), the nth term \( F_n \) in the Fibonacci series is the sum of the previous two terms
\begin{equation*}
F_n = F_{n-2} + F_{n-1}.
\end{equation*}
} % definition
We can quickly find that the series has values \( 0, 1, 1, 2, 3, 5, 8, 13, \cdots \).  What's really cool, is that there's a closed form expression for the nth term in the series that doesn't require calculation of all the previous terms.
\maketheorem{Nth term of the Fibonacci series.}{thm:fibonacci:40}{
\begin{equation*}
F_n = \frac{ \lr{ 1 + \sqrt{5} }^n - \lr{ 1 - \sqrt{5} }^n }{ 2^n \sqrt{5} }.
\end{equation*}
} % theorem
This is a rather miraculous and interesting looking equation.  Other than the \(\sqrt{5}\) scale factor, this is exactly the difference of the nth powers of the golden ratio \( \phi = (1+\sqrt{5})/2 \), and \( 1 - \phi = (1-\sqrt{5})/2 \).  That is:
\begin{dmath}\label{eqn:fibonacci:60}
F_n = \frac{\phi^n - (1 -\phi)^n}{\sqrt{5}}.
\end{dmath}

How on Earth would somebody figure this out?
% https://books.google.com/books?id=QGgLbf2oFUYC&pg=PA29
According to \href{https://books.google.com/books?id=QGgLbf2oFUYC&pg=PA29}{Tattersal} \citep{tattersall2005elementary}, this relationship was discovered by Kepler.

Understanding this from the ground up looks like it's a pretty deep rabbit hole to dive into.  Let's save that game for another day, but try the more pedestrian task of proving that this formula works.
\begin{proof}
\begin{dmath}\label{eqn:fibonacci:80}
\sqrt{5} F_n =
\sqrt{5} \lr{ F_{n-2} + F_{n-1} }
=
\phi^{n-2} - \lr{ 1 - \phi}^{n-2}
+ \phi^{n-1} - \lr{ 1 - \phi}^{n-1}
=
\phi^{n-2} \lr{ 1 + \phi }
-\lr{1 - \phi}^{n-2} \lr{ 1 + 1 - \phi }
=
\phi^{n-2}
\frac{ 3 + \sqrt{5} }{2}
-\lr{1 - \phi}^{n-2}
\frac{ 3 - \sqrt{5} }{2}.
\end{dmath}
However,
\begin{dmath}\label{eqn:fibonacci:100}
\phi^2
= \lr{ \frac{ 1 + \sqrt{5} }{2} }^2
= \frac{ 1 + 2 \sqrt{5} + 5 }{4}
= \frac{ 3 + \sqrt{5} }{2},
\end{dmath}
and
\begin{dmath}\label{eqn:fibonacci:120}
(1-\phi)^2
= \lr{ \frac{ 1 - \sqrt{5} }{2} }^2
= \frac{ 1 - 2 \sqrt{5} + 5 }{4}
= \frac{ 3 - \sqrt{5} }{2},
\end{dmath}
so
\begin{dmath}\label{eqn:fibonacci:140}
\sqrt{5} F_n = \phi^n - (1-\phi)^n.
\end{dmath}
\end{proof}
\section{How the square root fives cancel out.}
One of the interesting things in this Fibonacci formula, is the \( \sqrt{5} \)'s that are all over the place, while the formula represents only integer values.  Expanding the formula in binomial series shows us exactly why those terms all vanish.  Consider the first few values of \( n \) explicitly.
\begin{dmath}\label{eqn:fibonacci:160}
F_1
= \frac{ 1 + \sqrt{5} - \lr{ 1 - \sqrt{5} } }{ 2^1 \sqrt{5} }
= \frac{ 2 \sqrt{5} }{ 2^1 \sqrt{5} }
= 1,
\end{dmath}
\begin{dmath}\label{eqn:fibonacci:180}
F_2
= \frac{ 1 + 2 \sqrt{5} + 5 - \lr{ 1 - 2 \sqrt{5} + 5 } }{ 2^2 \sqrt{5} }
= \frac{ 4 \sqrt{5} }{ 2^2 \sqrt{5} }
= 1,
\end{dmath}
\begin{dmath}\label{eqn:fibonacci:200}
F_3
	= \frac{ 1 + 3 \sqrt{5} + 3 (5) + \sqrt{5} 5 - \lr{ 1 - 3 \sqrt{5} + 3(5) - \sqrt{5} 5 } }{ 2^3 \sqrt{5} }
= \frac{ 2 \lr{ 3 \sqrt{5} + \sqrt{5} 5 } }{ 2^3 \sqrt{5} }
= \frac{ 3 + 5 }{ 2^2 }
= 2.
\end{dmath}
In the general case, we have
\begin{dmath}\label{eqn:fibonacci:220}
2^n \sqrt{5} F_n
=
\sum_{k = 0}^n
\binom{n}{k}
{\sqrt{5}}^k
-
\sum_{k = 0}^n \binom{n}{k} (-\sqrt{5})^k
=
2 \sum_{1 \le k \le n, \mbox{$k$ is odd}} \binom{n}{k} (\sqrt{5})^k
=
2 \sqrt{5} \sum_{m = 0}^{\floor*{(n-1)/2}} \binom{n}{2 m + 1} 5^m,
\end{dmath}
% k = 2 m + 1, m = 0, 2 m + 1 <= n ; m <= (n-1)/2
so (for any \( n > 0 \)),
\begin{dmath}\label{eqn:fibonacci:240}
F_n =
\inv{2^{n-1}} \sum_{m = 0}^{\floor*{(n-1)/2}} \binom{n}{2 m + 1} 5^m.
\end{dmath}
Since only the odd powers of \( \sqrt{5} \) in the binomial expansions survive, the root in the basement is obliterated every time, leaving only integers upstairs, and a power of two factor downstairs.
It is still somewhat remarkable seeming that there is always a perfect cancellation of all the factors of two in the basement.
%, but we can see easily why the end result, for any \( n \) has no \( \sqrt{5} \) terms in it.
\section{Guessing the nth Fibonacci formula.}
We can rearrange the formula for the nth Fibonacci number as a difference equation
\begin{dmath}\label{eqn:fibonacci:260}
F_n - F_{n-1} = F_{n-2}.
\end{dmath}
This is a second order difference equation, so my naive expectation is that there are two particular solutions involved.  We know the answer, so it's not too hard to guess that the particular form of the solution has the following form
\begin{dmath}\label{eqn:fibonacci:280}
F_n = \alpha a^n + \beta b^n.
\end{dmath}
Given this guess, can we take some of the magic out of the formula, by just solving for \( \alpha, \beta, a, b \)?  Let's try that
\begin{equation}\label{eqn:fibonacci:300}
F_0 = \alpha + \beta = 0,
\end{equation}
\begin{dmath}\label{eqn:fibonacci:320}
F_1 = \alpha a + \beta b
    = \alpha \lr{ a - b } = 1,
\end{dmath}
and
\begin{dmath}\label{eqn:fibonacci:340}
F_n
= F_{n-1} + F_{n-2}
=
\alpha \lr{ a^{n-1} + a^{n-2} }
-\alpha \lr{ b^{n-1} + b^{n-2} }
=
\alpha a^{n-2} \lr{ 1 + a }
-\alpha b^{n-2} \lr{ 1 + b },
\end{dmath}
so
\begin{dmath}\label{eqn:fibonacci:360}
\begin{aligned}
a^2 &= a + 1 \\
b^2 &= b + 1.
\end{aligned}
\end{dmath}
If we complete the square we find
\begin{dmath}\label{eqn:fibonacci:380}
\lr{ a - \inv{2} }^2 = 1 + \inv{4} = \frac{5}{4},
\end{dmath}
or
\begin{dmath}\label{eqn:fibonacci:400}
a, b = \inv{2} \pm \frac{\sqrt{5}}{2}.
\end{dmath}
Out pop the golden ratio and it's complement.  Clearly we need to pick alternate roots for \( a \) and \( b \) or else we'd have zero for every value of \( n > 0 \).  Suppose we pick the positive root for \( a \), then to find the scaling constant \( \alpha \), we just compute
\begin{dmath}\label{eqn:fibonacci:420}
1 =
\alpha \lr{ \frac{ 1 + \sqrt{5}}{2} - \frac{ 1 - \sqrt{5} }{2} }
= \alpha \sqrt{5},
\end{dmath}
so our system \cref{eqn:fibonacci:280} has the solution:
\begin{equation}\label{eqn:fibonacci:520}
\begin{aligned}
a &= \frac{1 + \sqrt{5}}{2} \\
b &= \frac{1 - \sqrt{5}}{2} \\
\alpha &= \inv{\sqrt{5}} \\
\beta &= -\inv{\sqrt{5}}.
\end{aligned}
\end{equation}
It would be interesting to study difference equations enough to understand why the guess \cref{eqn:fibonacci:280} works.  However, we see a path that will systematically lead us from the Fibonacci difference equation to the final result, and have only to fill in a few missing steps to understand that path.

\paragraph{Another approach.}
I showed this to Sofia, and she came up with a neat very direct way to motivate the \( \sqrt{5} \).  It follows naturally (again knowing the answer), by assuming the Fibonacci formula has the following form:
\begin{dmath}\label{eqn:fibonacci:440}
F_n = \inv{x} \lr{
   \lr{ \frac{1 + x}{2}}^n
   -
   \lr{ \frac{1 - x}{2}}^n
}.
\end{dmath}
We have only to plug in \( n = 3 \) to find
\begin{dmath}\label{eqn:fibonacci:460}
2 x = \inv{4} \lr{ 1 + 3 x + 3 x^2 + x^3 - \lr{ 1 - 3 x + 3 x^2 - x^3 } }
= \inv{2} \lr{ 3 x + x^3 },
\end{dmath}
or
\begin{dmath}\label{eqn:fibonacci:480}
8 = 3 + x^2,
\end{dmath}
so
\begin{dmath}\label{eqn:fibonacci:500}
x = \pm \sqrt{5}.
\end{dmath}
Again the \( \sqrt{5} \)'s pop out naturally, taking away some of the mystery of the cool formula.
\section{Deriving the nth Fibonacci formula.}
There was a particularly unsatisfactory aspect of the \cref{eqn:fibonacci:280} guess.  In particular, we didn't have any reason to guess the form of that solution, except for the fact that we already knew the answer.  Now we will attempt to attack this in a more systematic fashion, so that each step along the way seems logical.  First, we need to put a couple goodies in our toolbox.
\makedefinition{Discrete sum.}{dfn:fibonacci:980}{
Given a set of discrete values \( \setlr{G_a, G_{a+1}, \cdots, G_n } \) we define a discrete sum of \( n - a + 1 \) of these terms as \( F_n \)
\begin{equation*}
F_n = \sum_{k = a}^n G_k + C,
\end{equation*}
where \( C \) is an arbitrary boundary value constant.
} % definition
\makedefinition{Difference operators.}{dfn:fibonacci:540}{
Define a backwards difference operator \( \Delta \), operating on \( X_n \) as
\begin{equation*}
\Delta X_n = X_n - X_{n-1}.
\end{equation*}
} % definition
The difference operator is a discrete analogue of a differential operator.  It is also possible to define a (forward) difference operator as \( \Delta X_n = X_{n+1} - X_{n} \), but the choice is arbitary, and we can find the same results either way.

\makelemma{Antidifference of discrete sum.}{thm:fibonacci:940}{
Given a sum \( F_n \) of the form \cref{dfn:fibonacci:980}, the difference operation is just the highest \( n \) term of the sum
\begin{equation*}
\Delta F_n = G_n.
\end{equation*}
} % theorem
\begin{proof}
\begin{dmath}\label{eqn:fibonacci:960}
\Delta F_n =
\sum_{k = a}^n G_k + C
- \lr{ \sum_{k = a}^{n-1} G_k + C }
= G_n.
\end{dmath}
\end{proof}

Computing differences is pretty easy.  What we want to do is the inverse operation (analogous to integration), where we find a closed form representation of \( F_n \) given a difference equation \( \Delta F_n = G_n \).
Just as we can compute antiderivatives for \( x^n \), we may do the same for \( n^k \) antidifferences, but the results are messier.  The first few such antidifferences are
\maketheorem{Antidifferences for powers of \(n\).}{thm:fibonacci:980}{
\begin{equation*}
\begin{aligned}
1   &= \Delta n \\
n   &= \Delta \lr{ \frac{n^2 + n}{2} } \\
n^2 &= \Delta \lr{ \frac{2 n^3 + 3 n^2 + n}{6} }.
\end{aligned}
\end{equation*}
} % theorem
\begin{proof}
Let's look at some example difference equations.  One of the simplest is probably a constant difference, such as
\begin{dmath}\label{eqn:fibonacci:560}
\Delta F_n = 1.
\end{dmath}
Like a differential equation, this has many possible solutions, so we need to fix initial conditions to find a specific solution.  For example, if we pick an initial value of \( F_0 = 0 \), the resulting sequence is
\begin{equation}\label{eqn:fibonacci:580}
\begin{aligned}
F_0 &= 0 \\
F_1 &= F_0 + 1 = 1 \\
F_2 &= F_1 + 1 = 2,
\end{aligned}
\end{equation}
and so forth.  By inspection that our system has the solution
\begin{dmath}\label{eqn:fibonacci:600}
F_n = n.
\end{dmath}
This can be generalized for arbitrary initial conditions, by simply adding an ``integration'' constant, so the general solution of \cref{eqn:fibonacci:560} is
\begin{dmath}\label{eqn:fibonacci:620}
F_n = n + C.
\end{dmath}
We can check the result by applying the difference operator
\begin{dmath}\label{eqn:fibonacci:640}
\Delta F_n = (n + C) - (n-1 + C) = 1,
\end{dmath}
as desired.

Let's try the difference operator on increasing powers of \( n \).  Starting with \( F_n = n^2 \) we have
\begin{dmath}\label{eqn:fibonacci:660}
\Delta n^2 = n^2 - (n-1)^2 = n^2 - (n^2 - 2 n + 1) = 2 n - 1,
\end{dmath}
and for \( F_n = n^3 \), we have
\begin{dmath}\label{eqn:fibonacci:680}
\Delta n^3
= n^3 - (n-1)^3
= n^3 - (n^3 - 3 n^2 + 3 n - 1)
= 3 n^2 - 3 n + 1.
\end{dmath}
We can work backwards from these to find a couple antidifference formulas
\begin{dmath}\label{eqn:fibonacci:700}
\Delta n^2 = 2 n - 1 = 2 n - \Delta n,
\end{dmath}
or
\begin{dmath}\label{eqn:fibonacci:720}
\Delta \frac{n^2 + n}{2} = n,
\end{dmath}
and
\begin{dmath}\label{eqn:fibonacci:740}
\Delta n^3
= 3 n^2 - 3 \Delta \frac{n^2 + n}{2}
+ \Delta n.
\end{dmath}
We can regroup
\begin{dmath}\label{eqn:fibonacci:760}
\Delta \frac{2 n^3 + 3 n^2 + n}{6} = n^2.
\end{dmath}
\end{proof}

This is now enough to find some sums that we have to use clever tricks to find otherwise.  For example
\begin{dmath}\label{eqn:fibonacci:780}
\Delta F_n = n,
\end{dmath}
written out explicitly is
\begin{dmath}\label{eqn:fibonacci:800}
F_n = F_{n-1} + n.
\end{dmath}
It's clear that nth term is just the sum of all the preceding values of \( n \), or
\begin{dmath}\label{eqn:fibonacci:820}
F_n = n + (n-1) + (n-2) + \cdots 1 + F_0.
\end{dmath}
Picking \( F_0 = 0 \) as the boundary conditions, we have
\begin{dmath}\label{eqn:fibonacci:840}
F_n = \sum_{k = 0}^n k,
\end{dmath}
so we can sum this using \cref{eqn:fibonacci:720}, to find
\begin{dmath}\label{eqn:fibonacci:860}
F_n = \frac{n^2 + n}{2} + C.
\end{dmath}
For \( F_0 = 0 \) we have \( C = 0 \), so
\begin{dmath}\label{eqn:fibonacci:880}
\sum_{k = 0}^n k = \frac{n}{2}(n+1).
\end{dmath}
This is the result that we know from the (probably apocryphal) story of Gauss's schoolboy summation of all the numbers from one to 99.  Know we know how to evaluate other sums.  For example:
\begin{dmath}\label{eqn:fibonacci:900}
F_n = \sum_{k = 0}^n k^2,
\end{dmath}
with \( F_0 = 0 \), using \cref{eqn:fibonacci:760}, we should have
\begin{dmath}\label{eqn:fibonacci:920}
\sum_{k = 1}^n k^2 = \frac{n}{6} \lr{ 2 n^2 + 3 n + 1 }.
\end{dmath}
Observe that no integration constant was required to find this sum of squares formula, because the chosen \( F_0 = 0 \) boundary condition was automatically satisfied.

Clearly we could do this for sums of any fixed power (sums of cubes, ...), but let's get back to the Fibonacci problem, now that we have some basic comfort with the ideas of difference equations.
%}
\EndArticle
